Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package

### 2021-07-07 doc/disbayes.Rmd
Restore doc dir with built vignettes

---
title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package}
 %\VignetteEngine{knitr::rmarkdown}
 \usepackage[utf8]{inputenc}
bibliography: disbayes.bib
---

@@

@@

If we assume that mortality from other causes is independent of disease status, deaths from other causes are uninformative and can be ignored.   Assume also for simplicity that remission from the disease is not possible (though the framework illustrated here is easily extensible to handle remission).

@@

* disease incidence $i(t)$, and the

* case fatality rate $f(t)$,

which are both assumed to depend on age $t$.  Assume further that these rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.

From these we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$. The matrix $P_a$ is defined as a function of $i_a$ and $f_a$, which is the solution to a differential equation, and is written out explicitly in the DisMod II paper (@dismod2).

@@

The next section discusses how these data can be derived, given the typical forms of data available from burden of disease studies.   A denominator is required because it enables uncertainty associated with each estimate to be quantified, as we will discuss later.

@@

This approach is used in the DisMod-MR software, as explained by @flaxman2015, however the software itself is undocumented and not fully published.  The older (published) DisMod II (@dismod2) used an ad-hoc optimisation approach to estimate parameters.

@@

* the ease of including multiple sources of direct/indirect data.  This is enabled by the computational methods and available software for Bayesian modelling, illustrated below.   This allows the approach to generalise to settings with different forms of data available.  In contrast, DisMod II only allows limited forms of data.

@@

The use of general-purpose free software for this computation, through the `disbayes` package, is described below.

# Converting data into the required form for the Bayesian model

@@

The following data are given in the R data frame `ihdbristol` supplied in the package, obtained from the Global Burden of Disease (REF) in a typical format:

* `inc` estimates of incidence of IHD by age

* `mort` estimates of IHD-specific mortality by age

* `pop` estimates of the population by age

and we wish to estimate case fatality by age, given these inputs.  A selection of rows from the full data frame are shown here.

```{r,show=FALSE}
library(disbayes)
ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
```

Note that the published "data" `inc` and `mort` are themselves estimates, based on underlying data from a finite population. To use the Bayesian approach, we need to recreate the _numerator_ and _denominator_ underlying each of these estimates.   We assume that

* the incidence has been estimated as $r_a^{(inc)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(inc)}$ of these get IHD before age $a+1$.

* the mortality has been estimated as $r_a^{(mort)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(mort)}$ of these die of IHD before age $a+1$.  These ratios are estimates of the true, underlying disease-specific mortalities $dm_a$.

In this example, $n_a$ and $r_a$ were not provided in the original data source, but can be roughly recreated, as follows.  We were given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5 (`pop` in the R data).

The number of incident IHD cases $r_a^{(inc)}$ can then be reconstructed by multiplying $n_a$ by the incidence rates `inc` supplied in the data.   Likewise, the number of IHD deaths $r_a$ (`ndieddis` in the R data) is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.

The data also contain estimates of prevalence, implicitly based on observing the number of people with IHD $r^{(prev)}_a$ in a sample $n^{(prev)}_a$.    The assumed sample size $n^{(prev)}_a$ underlying these estimates is given in the variable `prevdenom`, which is multiplied by the prevalence estimates (given as `prev`) to obtain estimated prevalence counts $r^{(prev)}_a$ (given as `prevn`).   Typically, prevalence is estimated from survey data, though for diseases with comprehensive registers, such as cancer, the denominator underlying the estimate might be assumed to equal the population size.

```{r,eval=TRUE}
ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
```

## ...Given point and interval estimates

Sometimes a point estimate $\hat{p}$ for a quantity such as incidence is published alongside an (e.g. 95%) interval estimate $(\hat{p}^{(lower)},\hat{p}^{(upper)})$.   The interval estimate can be assumed to express the uncertainty associated with the estimate.   Such information can be converted to an implicit numerator $r$ and denominator $n$  as follows.   We assume the point and interval estimate are summaries of a Beta posterior distribution which has been obtained by combining a vague prior with an observation of $r$ events occurring out of a sample of $n$ individuals.   If a Beta(0,0) prior is used, then the posterior is known to be Beta(r, n-r).  We can then search for the best-fitting Beta(r,n-r)$ distribution which has median $\hat{p}$ and (2.5,97.5) quantiles $(\hat{p}^{(lower)},\hat{p}^{(upper)})$, and set $r=a, n=a+b$.   A utility to perform this search is provided in the SHELF package for expert elicitation (REF).

## ...Expressing additional uncertainty

The uncertainty inherent in the information supplied about each of incidence, prevalence and mortality is measured by the denominator.  In the example above, this is the just size of the population used to produce the estimate.   However, if we also suspected that one of the data sources may be biased, but were unsure about the direction of bias, we could downweight that data source by multiplying both the numerator and denominator by the same amount, e.g. 0.5 if we wanted to give a data source half its original weight.  Note that if counts are scaled in this way, they should then be rounded to the nearest integer.

@@

	* (incidence) $r_a^{(inc)} \sim ~ Binomial(n_a, P_{a12})$, where $P_{a12}$ is the annual transition probability from no disease to disease.

@@

3. define prior distributions for the unknown parameters.  For incidence, currently the `disbayes` package assumes an unbounded uniform prior for each age independently.  For case fatality, the package allows two alternatives, explained below.

@@

In these examples, the case fatality is only informed indirectly.  The `disbayes` package implements two alternative models for how case fatality depends on age.

@@

2. the case fatality rate is assumed to be a smooth function $g()$ of age, $f_a = g(a)$.   This smooth function is defined by a spline basis $log(f_a) = \sum_k \beta_k g_k(a)$, where $g_k()$ are basis functions.  A "thin plate" spline is used, following the default in the `mgcv` package (Wood), and the amount of smoothness is determined automatically through a hierarchical prior on the coefficients (REF Wood, `jagam` paper).

@@

When running `disbayes`, both models are fitted by default.  Model (2) is more realistic.  However it may not give sensible estimates if the information provided by the data is too weak. The results of fitting Model (1) can help to diagnose where the indirect information on case fatality provided by the rest of the data is weaker or stronger.   If the information on $f_a$ for a particular age is too weak, then the posterior distribution will be identical to the prior.   In those cases, substantive prior information about case fatality at that age is necessary for the estimates to be meaningful.

@@

1. estimate with denominator: through arguments `inc` and `inc_denom`.  The numerator is computed automatically, as described above.

2. numerator and denominator, through arguments `inc_num` and `inc_denom`

@@

```{r, eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
dbres <- disbayes(dat = ihdbristol,
                 inc = "inc", 
                 inc_denom = "pop", 
                 prev_num = "prevn", 
                 prev_denom = "prevdenom",
                 mort = "mort",
                 mort_denom = "pop",
                 eqage = 40
                 )
```

The computation may take a few minutes, or less depending on the number of cores and their speed.

The `disbayes` function returns a list with components `fit` and `fitu`, containing the smoothed and unsmoothed model fits respectively.  Each of these components is an object in the `stanfit` format defined by the `rstan` R interface to Stan

@@

The results could be summarised and manipulated using functions provided by rstan, e.g. the `summary` and `extract` method for `stanfit` objects.    For example, we can check the MCMC simulations have converged, by examining the "trace plot" of the simulation progress for selected parameters. The simulated chains should mix together and look like white noise.

@@

The `disbayes` package provides some shortcuts for extracting summary statistics.   The posterior medians and 95% credible intervals can be extracted from the fitted smooth model using the `summary` method for `disbayes` objects.

The result is a data frame with the variables indicated in the row names.   These variables include `inc` (incidence), `cf` (case fatality), `mort` (mortality), `prev` (prevalence), each indexed by year of age.   The first few rows of the summary below contain estimates of incidence for the first few years of age.

```{r, eval=TRUE}
summ <- summary(dbres)
head(summ)
```

To extract results for a specific variable, the `vars` argument can be provided, e.g. for case fatality:

```{r,eval=TRUE}
summ <- summary(dbres, vars="cf")
head(summ)
```

To extract results from the unsmoothed model, the `summary_disbayes_fit` method is used.

```{r,eval=TRUE}
summ <- summary_disbayes_fit(dbres$fit, vars="cf")
summu <- summary_disbayes_fit(dbres$fitu, vars="cf")
head(summu)
```

The default `plot` method for objects returned by `disbayes` plots the posterior summaries for case fatality.  Both the smoothed and unsmoothed estimates from age 40 are shown.   This plot also overlays the corresponding estimates from DisMod II are overlaid in purple, included in the `ihdbristol` data in variable `dismod_cf`.

```{r, eval=TRUE, warning=FALSE}
library(ggplot2)
plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
```

@@

The smoothed model givese more precise estimates, aided by the assumption that case fatalities are similar between similar ages, and that case fatality is constant under the age of 40.  The smoothed model also removes the artefacts from the 5-year age grouping that are seen in the unsmoothed results.

The smoothed model agrees with the DisMod II estimates, and has the benefit of characterising uncertainty as a posterior distribution.  Uncertainty about case fatality in this example arises from the disease prevalence.   If the prevalence is lower, fewer people have the disease, giving less information from which to estimate the case fatality for people with the disease.

## Other things that could be done

* More ways of supplying substantive prior information

* Account for variations between populations in different areas.  Hierarchical modelling might be used here - DISMOD-MR does this, I believe.

* Hierarchical modelling to account for variations between ordered socioeconomic categories / income groups

* Borrow information between men/women

* Jointly model multiple diseases.  i.e. relaxing the assumption that mortality from causes other than the disease is the same for people with and without the disease.  Multimorbidity.

* Account for variations in incidence and prevalence with time as well as age.  DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.  Relaxing this assumption would be feasible, but possibly computationally expensive. 
	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011?

* In general, to transparently represent uncertainty, and explicitly build in other sources of information, in situations where we suspect the data may not be representative

All of these would be theoretically feasible in the above framework. The limits will be computational, and difficult to tell before getting hands on.

>>>>

---
title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 \usepackage[utf8]{inputenc}
 %\VignetteIndexEntry{Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package}
 %\VignetteEngine{knitr::rmarkdown}
bibliography: disbayes.bib
editor_options: 
  chunk_output_type: console
---

@@

A brief overview is given in this vignette, but a fuller explanation will be given in the paper _Bayesian multi-state modelling of incomplete chronic disease epidemiology data for health impact simulation models_ (Jackson et al., in progress).

@@

If we assume that mortality from other causes is independent of disease status, deaths from other causes are uninformative and can be ignored.

@@

* disease incidence $i(a)$, and the

* case fatality rate $f(a)$,

which are both assumed to depend on age $a$.  Remission is also possible for some diseases, with rate $r(a)$, though for simplicity of illustration is ignored here.  Assume further that the rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.

From these, we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$. The matrix $P_a$ is defined as a function of $i_a$ and $f_a$, which is the solution to a differential equation, and is written out explicitly in the DisMod II paper (@dismod2).

@@

The next section discusses how these data can be derived, given the typical forms of data available from burden of disease studies.   A denominator is required, because it enables uncertainty associated with each estimate to be quantified, as we will discuss later.

@@

This approach is used in the DisMod-MR package, as explained by @flaxman2015, however the software itself is not fully documented.  The older (published) DisMod II (@dismod2) used an ad-hoc optimisation approach to estimate parameters.

@@

@@

* the ease of including multiple sources of direct/indirect data.  This is enabled by the generic computational methods and available software for Bayesian modelling, specifically the Stan package, illustrated below.   This allows the approach to generalise to settings with different forms of data available.  Currently implemented models include hierarchical models, models with additive area and gender effects, and models with age-specific time trends.  In contrast, DisMod II only allows limited forms of data.

Further discussion of the differences between `disbayes` and DisMod-MR are given in the paper [Jackson et al, to appear].

# Data required by the Bayesian model

@@

The following data are given in the R data frame `ihdengland` supplied in the package.

* `inc_num`,`inc_denom`: incidence of IHD by age

* `mort_num`,`mort_denom`: estimates of IHD-specific mortality by age

* `prev_num`,`prev_denom`: estimates of prevalence by age

and we wish to estimate case fatality by age, given these inputs.  A selection of rows from the full data frame are shown here.  We select only the data from one area and gender (Bristol, male) for this illustration.

```{r,show=FALSE}
library(disbayes)
library(dplyr)
ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
ihdbristol %>% filter(between(age, 50, 55))
```

These data were obtained from the Global Burden of Disease, 2017, and transformed in three ways to obtain the data seen here.   Firstly, the published point estimates and confidence intervals for each annual risk probability was converted to an implicit numerator and denominator.  This is done by assuming that

* the incidence has been estimated as $r_a^{(inc)}/n_a^{(inc)}$, where out of $n_a^{(inc)}$ people who are alive at age $a$, $r_a^{(inc)}$ of these get IHD before age $a+1$.

* the mortality has been estimated as $r_a^{(mort)}/n_a^{(mort)}$, where out of $n_a^{(mort)}$ people who are alive at age $a$, $r_a^{(mort)}$ of these die of IHD before age $a+1$.  These ratios are estimates of the true, underlying disease-specific mortalities $dm_a$.

* the prevalence has been estimated as $r_a^{(prev)}/n_a^{(prev)}$, where out of $n_a^{(prev)}$ people who are alive at age $a$, $r_a^{(prev)}$ of them have IHD.

The published estimates and uncertainty interval are then used to derive the implicit denominator and numerator, using the method described in the next subsection.

Then secondly, estimates for five-year age groups were converted to one-year groups.  Most simply this could be done by assuming the counts are equal for each year of age, but a more sophisticated method was used based on "temporal disaggregation" using one of the methods implemented by Sax and Steiner in the `tempdisagg` R package (@RJ-2013-028).  This ensures that the disaggregated counts vary smoothly with age, while preserving the aggregate totals.   Thirdly, counts from smaller geographical areas were simply added up to obtain the corresponding counts for the larger areas (English city regions) required in this analysis.

```{r,eval=TRUE}
ihdbristol[ihdbristol$age %in% 50:55, ]
```

## Determining numerators and denominators from point and/or interval estimates

Sometimes a point estimate $\hat{p}$ for a quantity such as incidence is published alongside an (e.g. 95%) interval estimate $(\hat{p}^{(lower)},\hat{p}^{(upper)})$.   The interval estimate can be assumed to express the uncertainty associated with the estimate.   Such information can be converted to an implicit numerator $r$ and denominator $n$  as follows.   We assume the point and interval estimate are summaries of a Beta posterior distribution which has been obtained by combining a vague prior with an observation of $r$ events occurring out of a sample of $n$ individuals.   If a Beta(0,0) prior is used, then the posterior is known to be Beta(r, n-r).  We can then search for the best-fitting Beta(r,n-r)$ distribution which has median $\hat{p}$ and (2.5,97.5) quantiles $(\hat{p}^{(lower)},\hat{p}^{(upper)})$, and set $r=a, n=a+b$.   A utility to perform this search is provided in the SHELF package for expert elicitation (@ohagan:elic).

In other applications, the denominators might be directly available, e.g. we might know that the prevalence was estimated from a survey of $r_a^{(prev)}$ people, or we might assume that the incidence and mortality were obtained from a comprehensive registry of the population, in which case the denominator would equal the population size (within the specific stratum, e.g. as defined by age and gender, that the estimate refers to).

The uncertainty inherent in the information supplied about each of incidence, prevalence and mortality is measured by the denominator.  In the example above, this is equated to the size of the population used to produce the estimate.   However, if we also suspected that one of the data sources may be biased, but were unsure about the direction of bias, we could downweight that data source by multiplying both the numerator and denominator by the same amount, e.g. 0.5 if we wanted to give a data source half its original weight.  Note that if counts are scaled in this way, they should then be rounded to the nearest integer.

@@

	* (incidence) $r_a^{(inc)} \sim ~ Binomial(n_a, 1 - P_{a11})$, where $P_{a11}$ is the annual probability of remaining free of the disease.

@@

3. define prior distributions for the unknown parameters, as explained in the next section.

@@

In these examples, the case fatality is only informed indirectly.  The `disbayes` package implements two alternative models for how case fatality depends on age.  The same methods can be used for how incidence depends on age.

@@

2. the case fatality rate is assumed to be a smooth function $g()$ of age, $f_a = g(a)$.   This smooth function is defined by a spline basis $log(f_a) = \sum_k \beta_k g_k(a)$, where $g_k()$ are basis functions.  A "thin plate" spline is used, following the default in the `mgcv` R package (Wood), and the amount of smoothness is determined automatically through a hierarchical prior on the coefficients.

@@

Model (2) is more realistic.  However it may not give sensible estimates if the information provided by the data is too weak. The results of fitting Model (1) can help to diagnose where the indirect information on case fatality provided by the rest of the data is weaker or stronger.   If the information on $f_a$ for a particular age is too weak, then the posterior distribution will be identical to the prior.   In those cases, substantive prior information about case fatality at that age is necessary for the estimates to be meaningful.

@@

1. risk probability estimate with denominator: through arguments `inc` and `inc_denom`.  The numerator is computed automatically, by multiplying the denominator by the risk estimate.

2. numerator and denominator, through arguments `inc_num` and `inc_denom`.  This is what is done in the example here.

@@

```{r, eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
dbres <- disbayes(dat = ihdbristol,
                  age = "age",
                 inc_num = "inc_num", 
                 inc_denom = "inc_denom", 
                 prev_num = "prev_num", 
                 prev_denom = "prev_denom",
                 mort_num = "mort_num",
                 mort_denom = "mort_denom",
                 eqage = 40,
                 chains = 2,
                 iter = 1000
                 )
```

The computation may take a few minutes, or less depending on the number of cores and their speed.   The number of iterations `iter` should be high enough that the results are not affected by Monte Carlo error.

The `disbayes` function returns a list with components `fit` and `fitu`, containing the smoothed and unsmoothed model fits respectively.  Each of these components is an object in the `stanfit` format defined by the `rstan` R interface to Stan.

@@

The results could be summarised and manipulated using functions provided by rstan, e.g. the `summary` and `extract` method for `stanfit` objects.    For example, we can check the MCMC simulations have converged, by examining the "trace plot" of the simulation progress for selected parameters. The simulated chains should mix together and look like white noise.   Also if the `disbayes` call resulted in many reports of "divergent transitions", then the chains are unlikely to have converged to a valid posterior.  If convergence hasn't occurred, then the model probably needs to be simplified (e.g. by increasing `eqage`), or the data are insufficient.

@@

The `disbayes` package provides some shortcuts for extracting summary statistics.   The posterior medians, 95% credible intervals and other statistics for all parameters of interest can be extracted from the fitted models using the `tidy` method for `disbayes` objects.

The result is a data frame with the variables indicated in the variable called `var`.  Variables include `inc` (incidence), `cf` (case fatality), `mort` (mortality), `prev` (prevalence), each indexed by year of age given in the `age` variable.   `model` indicates the smoothed or unsmoothed model.  The first few rows of the summary below contain estimates of incidence for the first few years of age.

```{r, eval=TRUE}
summ <- tidy(dbres)
head(summ)
```

To extract results for a specific variable and model, it is convenient to use standard tidyverse functions, e.g. to get the median and interquartile range of case fatality under the smoothed model for people aged between 60 and 65:

```{r,eval=TRUE}
library(dplyr)
summ %>% 
  filter(var=="cf", between(age,60,65), model=="smoothed") %>%
  select(age, `25%`, `50%`, `75%`)
```

The default `plot` method for objects returned by `disbayes` plots the posterior summaries for case fatality.  Both the smoothed and unsmoothed estimates from age 40 and above are shown.    If other styles of plot are desired, the tidy data frame output of the `tidy` method above is designed so that it can be passed to `ggplot2` without too much effort.

```{r, eval=TRUE, warning=FALSE}
library(ggplot2)
plot(dbres) +  ylab("Case fatality") + xlab("Age") + 
  coord_cartesian(xlim=c(40,100), ylim=c(0,0.5)) 
```

@@

The smoothed model givese more precise estimates, aided by the assumption that case fatalities are similar between similar ages, and that case fatality is constant under the age of 40.  The smoothed model also removes the artefacts from the 5-year age grouping that are seen in the unsmoothed results.   While the point estimates agree with those from DisMod II (not shown here), the Bayesian method has the benefit of characterising uncertainty as a posterior distribution.  Uncertainty about case fatality in this example arises from the disease prevalence.   If the prevalence is lower, fewer people have the disease, giving less information from which to estimate the case fatality for people with the disease.

# Advanced models

`disbayes` provides a selection of more advanced models for more complex data structures.   These are substantially more computationally intensive than the basic model shown in this vignette, typically taking over an hour for applications similar to the one shown here.   See the paper for further information about how they are defined mathematically, and their R help pages for further information on how to invoke the functions.

## Hierarchical models

`disbayes_hier` is used to fit a joint model for data from multiple areas.

## Hierarchical models with additive area and gender effects

`disbayes_hier` can also be used to fit a joint model for data from multiple areas and different genders.  It is assumed that the relative case fatality between women and men is the same for each area.   While this is labelled `gender` it can be used for any single binary predictor, though usually we will want to consider gender.

## Trends through time (non-hierarchical models only)

`disbayes` can include information about trends through calendar time, which can be age dependent.  The ratio of incidence between previous years and the current year (the one represented by the data) can be supplied as a matrix, with one row per year of age and one column per calendar year.   See the help page, arguments `inc_trend_cf` and `cf_trend_mat`.
